Haploid Selection, Sex Ratio Bias, and Transitions Between Sex-Determination Systems

### 2017-10-25 LaTeX/Turnover.tex
haploid tight linkage appendix section

We next calculate when $(A)$ and $(B)$ are locally stable for $r=0$. 
According to the `small parameter theory' \citep{Karlin:1972ab,Karlin:1972dq}, these stability properties are unaffected by small amounts of recombination between the SDR and \textbf{A} locus, although equilibrium frequencies may be slightly altered. 
For the $a$ allele to be stably fixed on the Y we need $\bar{w}_{Ya}^{\male} > \bar{w}_{YA}^{\male}$ where $\bar{w}_{Ya}^{\male} = w_{a}^{\male} \big[\hat{p}_X^\female(1-\alpha^\male_\Delta) w_{A}^{\female} w_{Aa}^{\male} + (1-\hat{p}_X^\female)w_{a}^{\female} w_{aa}^{\male} \big]$ and $\bar{w}_{YA}^{\male} = w_{A}^{\male} \big[ \hat{p}_X^\female w_{A}^{\female} w_{AA}^{\male} + (1-\hat{p}_X^\female)(1+\alpha^\male_\Delta) w_{a}^{\female} w_{Aa}^{\male} \big]$. 
That is, Y-$a$ haplotypes must have higher fitness than Y-$A$ haplotypes.  
Substituting in $\hat{p}_X^\female = \hat{p}_{X}^{\female}$ from above, fixation of the $a$ allele on the Y requires that $\gamma_{i}>0$ where $\gamma_{(A)}=w_{a}^{\male} \big[ (1-\alpha^\male_\Delta) w_{Aa}^{\male} \phi + w_{aa}^{\male} \psi \big]-w_{A}^{\male} \big[ w_{AA}^{\male} \phi + (1 + \alpha^{\male}_\Delta)w_{Aa}^{\male} \psi \big]$ for equilibrium $(A)$ and $\gamma_{(B)}=(1-\alpha^{\male}_\Delta)w_{a}^{\male}w_{Aa}^{\male}-w_{A}^{\male}w_{AA}^{\male}$ for equilibrium $(B)$.
Stability of a polymorphism on the X chromosome (equilibrium $A$) further requires that $\phi >0$ and $\psi >0$. 
Fixation of the $a$ allele on the X (equilibrium $B$) can be stable only if equilibrium $(A)$ is not, as it requires $\psi<0$ and $2w_{A}^{\female}w_{AA}^{\female}>(1-\alpha_\Delta^{\female})w_{a}^{\female}w_{Aa}^{\female}$ or just $4w_{A}^{\female}w_{AA}^{\female}<(1-\alpha_\Delta^{\female})w_{a}^{\female}w_{Aa}^{\female}$ (which also prevents $\psi>0$).

@@

%As with equilibrium (B), if both haplotypes can spread ($\lambda_{mA}>1$ and $\lambda_{ma}>1$) then the neo-W invades under any rate of recombination with the selected locus, $R\geq0$. 
%In addition, even when only one haplotype can spread (e.g., under purely sexually-antagonistic selection $\lambda_{mA}>1$ and $\lambda_{ma}<1$), neo-W invasion can still occur under modest rates of recombination between the novel sex-determining and selected loci.

@@

\begin{figure}[!h]
\centering
\centerline{\includegraphics[width=\linewidth]{Region_Plot_combined_FemaleDrive_Mike}}
\caption{
ABC, $\alpha_{\Delta}^\female=-0.08$
DEF, $\alpha_{\Delta}^\female=0.08$. $\lambda$s are increased for the haplotype that is favoured by female haploid selection, the stability conditions are also affected. 
}
\label{fig:regionFemaleDrive}
\end{figure}

>>>>

We next calculate when $(A)$ and $(B)$ are locally stable for $r=0$. 
According to the `small parameter theory' \citep{Karlin:1972ab,Karlin:1972dq}, these stability properties are unaffected by small amounts of recombination between the SDR and \textbf{A} locus, although equilibrium frequencies may be slightly altered. 
For the $a$ allele to be stably fixed on the Y we need $\bar{w}_{Ya}^{\male} > \bar{w}_{YA}^{\male}$ where $\bar{w}_{Ya}^{\male} = w_{a}^{\male} \big[\hat{p}_X^\female(1-\alpha^\male_\Delta) w_{A}^{\female} w_{Aa}^{\male} + (1-\hat{p}_X^\female)w_{a}^{\female} w_{aa}^{\male} \big]$ and $\bar{w}_{YA}^{\male} = w_{A}^{\male} \big[ \hat{p}_X^\female w_{A}^{\female} w_{AA}^{\male} + (1-\hat{p}_X^\female)(1+\alpha^\male_\Delta) w_{a}^{\female} w_{Aa}^{\male} \big]$. 
That is, Y-$a$ haplotypes must have higher fitness than Y-$A$ haplotypes.  
Substituting in $\hat{p}_X^\female$ from equation \eqref{eq:tightequil}, fixation of the $a$ allele on the Y requires that $\gamma_{i}>0$ where $\gamma_{(A)}=w_{a}^{\male} \big[ (1-\alpha^\male_\Delta) w_{Aa}^{\male} \phi + w_{aa}^{\male} \psi \big]-w_{A}^{\male} \big[ w_{AA}^{\male} \phi + (1 + \alpha^{\male}_\Delta)w_{Aa}^{\male} \psi \big]$ for equilibrium $(A)$ and $\gamma_{(B)}=(1-\alpha^{\male}_\Delta)w_{a}^{\male}w_{Aa}^{\male}-w_{A}^{\male}w_{AA}^{\male}$ for equilibrium $(B)$.
Stability of a polymorphism on the X chromosome (equilibrium $A$) further requires that $\phi >0$ and $\psi >0$. 
Fixation of the $a$ allele on the X (equilibrium $B$) can be stable only if equilibrium $(A)$ is not, as it requires $\psi<0$ and $2w_{A}^{\female}w_{AA}^{\female}>(1-\alpha_\Delta^{\female})w_{a}^{\female}w_{Aa}^{\female}$ or just $4w_{A}^{\female}w_{AA}^{\female}<(1-\alpha_\Delta^{\female})w_{a}^{\female}w_{Aa}^{\female}$ (which also prevents $\psi>0$).

@@

\subsubsection*{Tight Linkage and Haploid Selection}

Generally, haploid selection expands the scenarios under which neo-W alleles can spread. 
For example, when selection is sexually-antagonistic in diploids ($s^\female s^\male <0$ and $0<h^\Hermaphrodite<1$) an unlinked neo-W ($R=1/2$) cannot invade unless there is also haploid selection, c.f., Figure \ref{fig:SexAntagTighter} and Figure \ref{fig:SexAntagTighterMaleDrive}. 
Secondly, with haploid selection, overdominance ($w_{aa}^\Hermaphrodite<w_{Aa}^\Hermaphrodite$ \& $w_{AA}^\Hermaphrodite<w_{Aa}^\Hermaphrodite$) is not required for neo-W-$a$ haplotypes to spread ($\lambda_{ma}>1$), Figures \ref{fig:regionMaleDrive}-\ref{fig:regionFemaleGS}. 
Finally, haploid selection can maintain a polymorphism in the face of directional selection in male and female diploids (ploidally-antagonistic selection). 
When selection is ploidally-antagonistic, neo-W alleles often spread, for at least some values of $R$, Figure \ref{fig:regionPloidAntag}.

As discussed above, male haploid selection alters the sex ratio and the alleles carried by male gametes that female gametes pair with. 
Male haploid selection in favour of the $a$ allele ($\alpha_{\Delta}^\male<0$, $w_{A}^\male<w_{a}^\male$) generates male-biased sex ratios at equilibria (A) and (B), where Y-$a$ is fixed ($\hat{p}_{Y}^\male=0$). 
Male-biased sex-ratios facilitate the spread of neo-W-$A$ and neo-W-$a$ haplotypes (increasing $\lambda_{WA}$ and $\lambda_{Wa}$). 
Panels A-C in Figures \ref{fig:regionMaleDrive} and \ref{fig:regionMaleGS} show that neo-W haplotypes tend to spread for a wider range of parameters when sex ratios are male biased, compared to Figure \ref{fig:regionplots} without haploid selection. 
By contrast, male haploid selection in favour of the $A$ allele generates female-biased sex ratios and reduces $\lambda_{WA}$ and $\lambda_{Wa}$, as demonstrated by panels D-F in Figures \ref{fig:regionMaleDrive} and \ref{fig:regionMaleGS}.

Female haploid selection generates direct selection on the neo-W-$A$ and neo-W-$a$ haplotypes as they spread in females. 
Thus, female haploid selection in favour of the $a$ allele tends to increase $\lambda_{Wa}$ and decrease $\lambda_{WA}$, as shown by panels A-C in Figures \ref{fig:regionFemaleDrive} and \ref{fig:regionFemaleGS}. 
Conversely, female haploid selection in favour of the $A$ allele increases $\lambda_{WA}$ and decreases $\lambda_{Wa}$, see panels D-F in Figures \ref{fig:regionFemaleDrive} and \ref{fig:regionFemaleGS}.

@@

\begin{figure}[!h]
\centering
\centerline{\includegraphics[width=\linewidth]{Region_Plot_combined_FemaleDrive_Mike}}
\caption{
ABC, $\alpha_{\Delta}^\female=-0.16$
DEF, $\alpha_{\Delta}^\female=0.16$. $\lambda$s are increased for the haplotype that is favoured by female haploid selection, the stability conditions are also affected. 
}
\label{fig:regionFemaleDrive}
\end{figure}
